Reproducibility of SNP calling in multiple sequencing runs from single tumors

### 2015-03-20 PAPER/paper_d3.tex
not sure if there is anything

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

@@

@@

\title{Reproducibility of SNP calling in multiple sequencing runs from single tumors}
\author{Dakota Z. Derryberry, Claus O. Wilke, Matthew C. Cowperthwaite} %**ask them about author order, I'm out of it...

@@

The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from a single patient for thousands of tumors \cite{TCGA-GBM, TCGA-GBM-13}. TCGA's database includes hundreds of samples of each of several tumor types, including 580 and counting samples alone of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor. GBM has a median survival of 14 months and a 5-year survival of ~5\%, and prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotheraputics \cite{GBM-stats}. In an effort to improve patient outcomes, TCGA and other groups \cite{Parsons} have used large scale genomic data to discover which genes and pathways are mutated in GBM \cite{pathways}, discover different GBM subtypes \cite{subtypes}, and develop a variety of computational models to find GBM driver mutations \cite{drivers}.

% How the hell do I cite anything in this paragraph? There is NOTHING TO CITE. (Except that power point intro???)
Despite widespread use of TCGA data in these and other research efforts, and significant monetary investment in collecting the data, there has been very little investigation of the best way to sequence the 580+ paired tumor/normal samples in the TCGA GBM database. Most samples in the database include, at a minimum, whole exam sequencing of both the tumor and a matched blood sample at a coverage depth of between 8X and 80X, usually around 10--15X in the blood and 20--50X in the tumor. This is low, but similar to the 30X recommended for variant calling by the Broad Institute Introduction to NGS. However, we can find no research looking at the accuracy of variant calling at this or any sampling depth, for common or rare variants, in normal or cancerous tissue. Given that cancer is thought to be genetically heterogenous \cite{heterogenous1, heterogenous2}, and that rare variants in GBM may be functionally important \cite{rare}, we think that some inquiry into the accuracy and precision of standard cancer sequencing methods is warranted.

In this paper, we begin this inquiry with two simple questions: Given two technical sequencing replicates of a single tumor, how similar are they? Does SNP-calling software increase or decrease the degree similarity? We answered these questions using data from TCGA, which includes 55 GBM tumors that were sequenced twice, once each with (i) the standard WGS protocol, and (ii) an additional amplification step before library prep, which we refer to as the WGA protocol. For each of these 55 samples, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using SomaticSniper \cite{SomaticSniper}. We found significant overlap (around 50\%) between technical replicates, but also significant differences. As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates. Still, the number of mutations in the WGS replicates only varied by orders of magnitude, from under 100 to almost 10,000. Contrary to expectations, the quality of results (measured by the percentage overlap between technical replicates) did not decrease with increasing numbers of mutations, suggesting real variation in mutation frequency between samples. This may be in part due to know mutational hotspots in some tumors \cite{Karen}. Using SomaticSniper and some custom protocols to analyze the likelihood that individual variants were true positives (actual somatic mutations in the tumor, and not sequencing or analysis errors or various sorts) was unsuccessful, because almost no putative SNPs were identified as probably SNPs.

@@

\subsection{Just how similar are two iterations of sequencing the same tumor?}

@@

DNA sequencing is not error-free. Error is introduced by mis-called bases in sequencing runs, and by mis-aligned bases during sequence analysis. Loss of heterozygosity also occurs in sequencing, if only one allele of a polymorphic site is amplified. Cancer DNA is highly heterogenous, which makes for an additional source of error: a mutation present in only some tumor cells may or may not be present in a sample that is sequenced, so that a partially penetrant mutation may appear fixed or not at all. In general, we would like to know how often these error occur. One way of investigating would be to use PCR to verify each individual mutation, but this method is expensive and time consuming. A cheaper alternative would be to sequence the tumor multiple times, and look at the similarity between the iterations. Theoretically, any mutation fixed throughout the tumor should appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors would not. (Incompletely penetrant mutations would also not be fixed in all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) This is the method used in most biological sequencing experiments--but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research. Still, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know about what percentage of called mutations would or would not likely appear in additional sequencing replicates.

@@

For each patient (n=55), we called mutations in both technical replicates and the patient's blood sample using the same computational pipeline (see Figure 1 and Methods): we downloaded TCGA bamfiles with CGHub, re-generated fastq files with picard, re-aligned the fastq files to hg19 with bwa, performed indel re-alignment and base recalibration with GATK, and finally called somatic mutations with SomaticSniper. We then compared the lists of putative SNPs generated by the WGS and WGA bamfiles for each patient, and calculated the length of each list and the length of the overlap, or intersection, between the two lists (Figure 2). We further calculated the percent of the WGA list that overlapped with the WGS list and the percent of the WGS list that overlapped with the WGA list. We found that the number of mutations found in one replicate correlated strongly with the number of mutations found in the other, though as expected the WGA replicate, with the additional amplification step, had slightly more mutations (Figure 3). We further found that the percent overlap between the two samples, calculated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was fairly consistent, mostly between 20--40\% in WGA replicates and mostly between 40--60\% in WGS replicates. As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure 4).

% Experiment: number of mutations (figure 5)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
Although the percent overlap in WGS and WGA samples remained fairly constant, and the number of possible mutations in WGS and WGA lists was correlated, the exact number of putative mutations varied across samples by orders of magnitude (less than 100 to over 10,000). Different cancers mutate at different rates: some pediatric cancers that arise from single and double hit mutations \cite{pediatric}, while some tumors might have a mutator phenotype, which usually results from errors in DNA repair pathways \cite{mutator}. GBM specifically is thought to have a relatively low mutation rate \cite{GBM_mut_rate}, and while the majority of our samples had low mutation frequencies in line with this theory, several samples also had mutation frequencies an order of magnitude greater. One possible explanation is a degraded DNA sample, or bad data. If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations. We found no correlation between the number of putative mutations and the percentage of the sample that was overlapping (Figure 5). That the percentage overlap does not vary with the number of putative mutations in either replicate suggests that some samples may simply have a higher mutation rate than others, or indeed than is generally supposed in GBM.

\subsection{Does SNP-calling software increase or decrease the degree similarity between replicates?}

% Experiment: which filters do what (figures 4 and 5) % Experiment: those two filters are different (figures 6 and 7)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
The aim of sequencing tumors is generally to find somatic mutations, those mutations that arise in a tumor and are not present in the rest of the organism. To do this, researchers sequence a patient's tumor and blood and align the two to each other; Differences between the two sequences are putative somatic mutations. Because sequencing is not error free, and because a patient's blood tissue might also have somatic mutations different from the tumor, not all of the putative somatic mutations are mutations, and some method of distinguishing true somatic mutations from sequencing and other errors is needed. If possible, this is done by individually validating each putative mutation with PCR; however, due to restraints on time and money, as well as the amount of tumor tissue available and it's purity, this is often not possible. In this case, computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing error may be employed.

Software platforms to distinguish somatic mutations from germ line mutations and sequencing error are plentiful, and each one typically employs multiple methods to identify true positive mutations. SomaticSniper, the platform used in this research, uses several probabilistic calculations based on features of the dataset to sort out true putative mutations from false ones. Another example, the GATK QualityScore, calculates the probability that a mutation is a true mutations based on features of the reads. In this paper, we used eight different methods, or filters, to identify somatic mutations, summarized in Table 1. The first three are the quality scores generated by the GATK and SomaticSniper (we simply remove from the dataset anything that fails to meet these thresholds). The last five we coded ourselves in Python. Each of these five filters represents a quality of the data or putative mutation that is generally thought to indicate that it is probably a false positive.

@@

GBM, primary brain cancer, is an evolutionary process by which mutations arise in glial cell lines and these mutated cells and their lineages co-opt the surrounding tissue and systems to the detriment of the organism of the whole. Treatment for GBM is difficult and has poor outcomes \cite{GBM-treatments}, but could hopefully be improved by a more complete understanding of the somatic mutations present in GBM. Large-scale sequencing projects, like TCGA, make cancer sequencing data available to many researchers. This data is enormously valuable to the research community, but the accuracy and reproducibility of these data are unknown, and knowing them could only increase the utility and value of the data. We make a first pass at evaluating the accuracy and repeatability of the data by comparing 55 technical replicates in the TCGA GBM dataset.

@@

Filtering the raw computational data using both custom filters and those in the software SomaticSniper eliminated more than half of the total number of putative mutations in all 110 samples, including most or all of those present in both replicates.  Of the six filters whose individual effects we examined, only two, those that removed Loss of Heterozygosity (LOH) mutations and putative mutations with low VAQ (calculated by SomaticSniper), removed primarily mutations that found in both technical replicates. Combined, the two removed most or all of those mutations present in both replicates. Despite this, two still appeared to remove different sets of mutations. Of the remaining four, only three removed any appreciable number of putative mutations from the sample, and each of these preferentially removed mutations present in only one technical replicate. Two of these three filters, those removing putative mutations within a 10 bp window of other putative mutations or within a 10 bp window of a putative indel, recognized a feature (clustered mutations) that suggested a local problem with the reads or alignment, and removed this data from consideration. The third removed overlap with dbGaP.  Our analysis suggests that these three filters are useful, and do clean up the purely computational data in a meaningful way. By contrast, it may be more useful to discard the two filters that removed principally data from the overlap of the two replicates. Of these, one, VAQ is built into SomaticSniper. The second removes loss of heterozygosity (LOH) mutations. Multiple sources suggest that LOH mutations may be essential to cancer \cite{LOH}. This, along with the data presented here, makes a strong argument for retaining these mutations in functional analyses rather than excluding them.

Several factors limit the conclusions we may draw from this analysis. First, in this analysis we use repeatability between technical overlaps (being in the WGS and WGA samples) as a measure of confidence in a putative SNP. This metric is potentially problematic for two reasons: (i) cancer is highly heterogeneous \cite{hetero}, and so a legitimate somatic SNP might show up in one replicate and not another; and (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates. Although repeatability does not and could not perfectly measure confidence, it should correlate. Having an independent measure of the confidence in SNP calls (repeatability), even an imperfect one, can help us gauge the accuracy of other measures (filtering).  In this sample, we have shown that while some filters remove primarily putative SNPs in only one of two replicates (more likely to be false positives), two filters, VAQ and LOH, remove most of the overlap (more likely to be true positives). While our results suggest that VAQ and LOH filters are probably removing some true positives, that does not preclude them removing false positives. Therefore, if we remove these filters, we may retain more true positives, but we also retain more false positives.

@@

We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNPs in whole exome sequencing and about 30\% in whole genome sequencing. This effect remains even at a high number of putative SNPs, suggesting that some GBMs may have significantly more somatic mutation than others. While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six strictly computational filters on the putative mutations in each technical replicate and the overlap between them, for all 55 samples. We found that some filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap. We suggest using just the first set in the future, when mutation validation by PCR is not an option.

@@

\subsection{Data and back-end processing}

All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \cite{TCGA-GBM}. We downloaded three bamfiles for each of 55 patients using CGHub \cite{CGHub}. For each patient, data consisted of one bamfile taken from blood DNA, and two bamfiles from tumor DNA, one for each technical replicate. In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library.

We created a pipeline for backend processing of all TCGA bamfiles, summarized in Figure 1, with commands given in Table 2. The custom python code to connect the pipeline is available in a public 
git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}). our pipeline first regenerated fastq files (original reads) from the TCGA bamfiles, which were aligned to hg18, using picard. Next, we used BWA to align the fastq files to hg19, and samtools to sort, index, clean, and de-duplicate the new bamfile. We used GATK to do indel realignment and base recalibration, according to the standards best practices for genomics data \cite{best-practices},. Finally, we predicted somatic variants with SomaticSniper \cite{SomaticSniper}, with the output given in VCF format.

\subsection{Filtering and data analysis}

@@

We used the same or substantially similar python scripts to calculate and compare the overlap between technical replicates before and after filtering. These scripts are also available in the public git repository. We plotted all data and did all statistics with standard plotting and statistics functions in R \cite{Rsoftware}. This code is also available in the public git repository.

\section*{}

\bibliography{TCGA_bib}
\bibliographystyle{plain}

>>>>

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customize the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

@@

%%%Bibliography
\usepackage{natbib}

@@

\title{Reproducibility of SNP calling in multiple sequencing runs from single tumors}
\author{Dakota Z. Derryberry, Matthew C. Cowperthwaite, Claus O. Wilke} %**ask them about author order, I'm out of it...

@@

The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from a single patient for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}. TCGA's database includes hundreds of samples of each of several tumor types, including 580 and counting samples alone of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor. GBM has a median survival of 14 months and a 5-year survival of ~5\%, and prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotheraputics \citep{GBM-stats}. In an effort to improve patient outcomes, TCGA and other groups \citep{Parsons} have used large scale genomic data to discover which genes and pathways are mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.

Despite widespread use of TCGA data in these and other research efforts, and significant monetary investment in collecting the data, there has been very little investigation of the best way to sequence the 580+ paired tumor/normal samples in the TCGA GBM database. Most samples in the database include, at a minimum, whole exam sequencing of both the tumor and a matched blood sample at a coverage depth of between 8X and 80X, usually around 10--15X in the blood and 20--50X in the tumor. This is low, but similar to the 30X recommended for variant calling by the Broad Institute Introduction to NGS. However, we can find no research looking at the accuracy of variant calling at this or any sampling depth, for common or rare variants, in normal or cancerous tissue. Given that cancer is thought to be genetically heterogenous \citep{heterogenous1, heterogenous2}, and that rare variants in GBM may be functionally important \citep{rare}, we think that some inquiry into the accuracy and precision of standard cancer sequencing methods is warranted.

In this paper, we begin this inquiry with two simple questions: Given two technical sequencing replicates of a single tumor, how similar are they? Does SNP-calling software increase or decrease the degree similarity? We answered these questions using data from TCGA, which includes 55 GBM tumors that were sequenced twice, once each with (i) the standard WGS protocol, and (ii) an additional amplification step before library prep, which we refer to as the WGA protocol. For each of these 55 samples, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using SomaticSniper \citep{SomaticSniper}. We found significant overlap (around 50\%) between technical replicates, but also significant differences. As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates. Still, the number of mutations in the WGS replicates only varied by orders of magnitude, from under 100 to almost 10,000. Contrary to expectations, the quality of results (measured by the percentage overlap between technical replicates) did not decrease with increasing numbers of mutations, suggesting real variation in mutation frequency between samples. This may be in part due to know mutational hotspots in some tumors \citep{Karen}. Using SomaticSniper and some custom protocols to analyze the likelihood that individual variants were true positives (actual somatic mutations in the tumor, and not sequencing or analysis errors or various sorts) was unsuccessful, because almost no putative SNPs were identified as probably SNPs.

@@

\subsection*{Just how similar are two iterations of sequencing the same tumor?}

@@

DNA sequencing is not error-free. Error is introduced by mis-called bases in sequencing runs, and by mis-aligned bases during sequence analysis \citep{seqerror}. Loss of heterozygosity also occurs in sequencing, if only one allele of a polymorphic site is amplified. Cancer DNA is highly heterogenous, which makes for an additional source of error: a mutation present in only some tumor cells may or may not be present in a sample that is sequenced, so that a partially penetrant mutation may appear fixed or not at all. In general, we would like to know how often these error occur. One way of investigating would be to use PCR to verify each individual mutation, but this method is expensive and time consuming. A cheaper alternative would be to sequence the tumor multiple times, and look at the similarity between the iterations. Theoretically, any mutation fixed throughout the tumor should appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors would not. (Incompletely penetrant mutations would also not be fixed in all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) This is the method used in most biological sequencing experiments--but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research. Still, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know about what percentage of called mutations would or would not likely appear in additional sequencing replicates.

@@

For each patient (n=55), we called mutations in both technical replicates and the patient's blood sample using the same computational pipeline (see Figure 1 and Methods): we downloaded TCGA bamfiles with CGHub \citep{CGHub}, re-generated fastq files with picard \citep{picard}, re-aligned the fastq files to hg19 with bwa \citep{bwa}, performed indel re-alignment and base recalibration with GATK \citep{GATK}, and finally called somatic mutations with SomaticSniper \citep{SomaticSniper}. We then compared the lists of putative SNPs generated by the WGS and WGA bamfiles for each patient, and calculated the length of each list and the length of the overlap, or intersection, between the two lists (Figure 2). We further calculated the percent of the WGA list that overlapped with the WGS list and the percent of the WGS list that overlapped with the WGA list. We found that the number of mutations found in one replicate correlated strongly with the number of mutations found in the other, though as expected the WGA replicate, with the additional amplification step, had slightly more mutations (Figure 3). We further found that the percent overlap between the two samples, calculated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was fairly consistent, mostly between 20--40\% in WGA replicates and mostly between 40--60\% in WGS replicates. As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure 4).

% Experiment: number of mutations (figure 5)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
Although the percent overlap in WGS and WGA samples remained fairly constant, and the number of possible mutations in WGS and WGA lists was correlated, the exact number of putative mutations varied across samples by orders of magnitude (less than 100 to over 10,000). Different cancers mutate at different rates: some pediatric cancers have very few mutations \citep{RB2hit, pediatric}; other tumors show a mutator phenotype, leading to vastly more mutations, usually resulting from errors in DNA repair pathways \citep{mutator}. GBM specifically is thought to have a relatively low mutation rate \citep{Parsons, TCGA-GBM-13}, and while the majority of our samples had low mutation frequencies in line with this theory, several samples also had mutation frequencies an order of magnitude greater. One possible explanation is a degraded DNA sample, or bad data. If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations. We found no correlation between the number of putative mutations and the percentage of the sample that was overlapping (Figure 5). That the percentage overlap does not vary with the number of putative mutations in either replicate suggests that some samples may simply have a higher mutation rate than others, or indeed than is generally supposed in GBM.

\subsection*{Does SNP-calling software increase or decrease the degree similarity between replicates?}

% Experiment: which filters do what (figures 4 and 5) % Experiment: those two filters are different (figures 6 and 7)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
The aim of sequencing tumors is generally to find somatic mutations, those mutations that arise in a tumor and are not present in the rest of the organism. To do this, researchers sequence a patient's tumor and blood and align the two to each other; Differences between the two sequences are putative somatic mutations. Because sequencing is not error free, and because a patient's blood tissue might also have somatic mutations different from the tumor, not all of the putative somatic mutations are mutations, and some method of distinguishing true somatic mutations from sequencing and other errors is needed. If possible, this is done by individually validating each putative mutation with PCR; however, due to restraints on time and money, as well as the amount of tumor tissue available and its purity, this is often not possible. In this case, computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing error may be employed \citep{SomaticSniper, mut_calling}.

Software platforms to distinguish somatic mutations from germ line mutations and sequencing error are plentiful, and each one typically employs multiple methods to identify true positive mutations. SomaticSniper, the platform used in this research, uses several probabilistic calculations based on features of the dataset to sort out true putative mutations from false ones \citep{SomaticSniper}. Another example, the GATK QualityScore, calculates the probability that a mutation is a true mutations based on features of the reads \cite{GATK}. In this paper, we used eight different methods, or filters, to identify somatic mutations, summarized in Table 1. The first three are the quality scores generated by the GATK and SomaticSniper (we simply remove from the dataset anything that fails to meet these thresholds). The last five we coded ourselves in Python. Each of these five filters represents a quality of the data or putative mutation that is generally thought to indicate that it is probably a false positive.

@@

GBM, primary brain cancer, is an evolutionary process by which mutations arise in glial cell lines and these mutated cells and their lineages co-opt the surrounding tissue and systems to the detriment of the organism of the whole. Treatment for GBM is difficult and has poor outcomes \citep{GBM-stats}, but could hopefully be improved by a more complete understanding of the somatic mutations present in GBM. Large-scale sequencing projects, like TCGA, make cancer sequencing data available to many researchers. This data is enormously valuable to the research community, but the accuracy and reproducibility of these data are unknown, and knowing them could only increase the utility and value of the data. We make a first pass at evaluating the accuracy and repeatability of the data by comparing 55 technical replicates in the TCGA GBM dataset.

@@

Filtering the raw computational data using both custom filters and those in the software SomaticSniper eliminated more than half of the total number of putative mutations in all 110 samples, including most or all of those present in both replicates.  Of the six filters whose individual effects we examined, only two, those that removed Loss of Heterozygosity (LOH) mutations and putative mutations with low VAQ (calculated by SomaticSniper), removed primarily mutations that found in both technical replicates. Combined, the two removed most or all of those mutations present in both replicates. Despite this, two still appeared to remove different sets of mutations. Of the remaining four, only three removed any appreciable number of putative mutations from the sample, and each of these preferentially removed mutations present in only one technical replicate. Two of these three filters, those removing putative mutations within a 10 bp window of other putative mutations or within a 10 bp window of a putative indel, recognized a feature (clustered mutations) that suggested a local problem with the reads or alignment, and removed this data from consideration. The third removed overlap with dbGaP.  Our analysis suggests that these three filters are useful, and do clean up the purely computational data in a meaningful way. By contrast, it may be more useful to discard the two filters that removed principally data from the overlap of the two replicates. Of these, one, VAQ is built into SomaticSniper. The second removes loss of heterozygosity (LOH) mutations. Multiple sources suggest that LOH mutations may be essential to cancer \citep{LOH}. This, along with the data presented here, makes a strong argument for retaining these mutations in functional analyses rather than excluding them.

Several factors limit the conclusions we may draw from this analysis. First, in this analysis we use repeatability between technical overlaps (being in the WGS and WGA samples) as a measure of confidence in a putative SNP. This metric is potentially problematic for three reasons: (i) cancer is highly heterogeneous, and so a legitimate somatic SNP might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; and (iii) a putative mutation that is in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}. Although repeatability does not and could not perfectly measure confidence that a putative mutation is a somatic mutation, it does make it more likely. Having an independent measure of the confidence in SNP calls (repeatability), even an imperfect one, can help us gauge the accuracy of other measures (filtering).  In this sample, we have shown that while some filters remove primarily putative SNPs in only one of two replicates (more likely to be false positives), two filters, VAQ and LOH, remove most of the overlap (more likely to be true positives). While our results suggest that VAQ and LOH filters are probably removing some true positives, that does not preclude them removing false positives. Therefore, if we remove these filters, we may retain more true positives, but we also retain more false positives.

@@

We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNPs in WGS samples and about 30\% in WGA samples. This effect remains even at a high number of putative SNPs, suggesting that some GBMs may have significantly more somatic mutation than others. While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six strictly computational filters on the putative mutations in each technical replicate and the overlap between them, for all 55 samples. We found that some filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap. We suggest using just the first set in the future, when mutation validation by PCR is not an option.

@@

\subsection*{Data and back-end processing}

All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \citep{TCGA-GBM}. We downloaded three bamfiles for each of 55 patients using CGHub \citep{CGHub}. For each patient, data consisted of one bamfile taken from blood DNA, and two bamfiles from tumor DNA, one for each technical replicate. In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library \citep{TCGA-GBM}.

We created a pipeline for backend processing of all TCGA bamfiles, summarized in Figure 1, with commands given in Table 2. The custom python code to connect the pipeline is available in a public 
git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}). Our pipeline first regenerated fastq files (original reads) from the TCGA bamfiles, which were aligned to hg18, using picard \citep{picard}. Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools}to sort, index, clean, and de-duplicate the new bamfile. We used GATK \citep{GATK} to do indel realignment and base recalibration, according to the standards best practices for genomics data \citep{best-practices}. Finally, we predicted somatic variants with SomaticSniper \citep{SomaticSniper}, with the output given in VCF format.

\subsection*{Filtering and data analysis}

@@

We used the same or substantially similar python scripts to calculate and compare the overlap between technical replicates before and after filtering. These scripts are also available in the public git repository. We plotted all data and did all statistics with standard plotting and statistics functions in R \citep{Rsoftware}. This code is also available in the public git repository.

%\section*{}

\bibliography{TCGA_bib}
\bibliographystyle{peerj}
